Retrofitting Effect Handlers onto OCaml

### 2020-11-02 retro-concurrency.tex
Implementation section is finished

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables to obtain a backtrace. Our primary motivation for adding effect handlers
to OCaml is to support (1) highly concurrent applications such as web servers
to be expressed naturally in \emph{direct-style} as opposed to using callbacks
and (2) efficient data and task parallel programs in the vein of Cilk~\cite{},
OpenMP~\cite{} and Intel TBB~\cite{}.

@@

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
  \begin{minipage}{0.39\linewidth}
    \centering
    \includegraphics[scale=0.46]{figures/stock_stack}
    \subcaption{Stack layout before raising exception.}
    \label{fig:stock_stack}
  \end{minipage}
  \caption{Program stack on stock OCaml.}
\end{figure*}

@@

@@

In order to illustrate the implementation of handling effects, consider the
example presented below:
\begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *) (*p1*) perform E (*p3*)
  with v -> v | effect F kf -> continue kf ()
with v -> v | effect E ke -> (*p2*) continue ke ()
\end{lstlisting}
The code performs |E| which is handled in the outer-most handler, and is
immediately resumed. The layout of the stack as the program executes is
captured in Figure~\ref{fig:fiber_handler}. At position |p1|, |rsp| is at the
top of the fiber |f|.

@@

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|).

@@

>>>>

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient data and task parallel programs in the vein
of Cilk~\cite{}, OpenMP~\cite{} and Intel TBB~\cite{}.

@@

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
  \begin{minipage}{0.39\linewidth}
    \centering
    \includegraphics[scale=0.46]{figures/stock_stack}
    \subcaption{Stack layout before raising exception \texttt{E1}.}
    \label{fig:stock_stack}
  \end{minipage}
  \caption{Program stack on stock OCaml.}
\end{figure*}

@@

\subsection{Stack unwinding}
\ref{sec:unwind}

@@

In order to illustrate the implementation of handling effects, consider the
example presented below:
\begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *) (*p1*) perform E (*p3*)
  with v -> v | effect F kf -> continue kf ()
with v -> v | effect E ke -> (*p2*) continue ke ()
\end{lstlisting}
The code performs |E| which is handled in the outer-most handler, and is
immediately resumed. The layout of the stack as the program executes is
captured in Figure~\ref{fig:fiber_handler}. The arrows between rectangular
fibers are parent pointers. At position |p1|, |rsp| is at the top of the fiber
|f|.

@@

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|). We perform standard tail-call optimization so that resumptions
at tail positions do not build up stack.

@@

For stack walking during GC, the only change necessary was to make the runtime
aware of fibers layout and how to follow the parent pointers. The stack map
generation and scanning each frame for roots remains the same. Due to the
uniformity of fiber layout, suspended fibers are scanned the same way as the
current active fiber. Previous work~\cite{RetroParallel} presents the
challenges and solution for the interaction of fibers and the concurrent
mark-and-sweep GC of Multicore OCaml.

The challenge with DWARF stack unwinding is to make it aware of the
non-contiguous stacks. While the complete details of DWARF stack unwinding is
beyond the scope of the paper, it is beneficial to know how DWARF unwind tables
are constructed in order to appreciate our solution. We refer the interested
reader to Bastian et al.~\cite{} for a good overview of DWARF stack unwinding.

Logically, DWARF maintains a large table which records for every machine
instruction where the return address and callee-saved registers are stored. To
avoid reifying this large table, DWARF represents the table using a compact,
turing-complete, bytecode representation, which is interpreted on demand for
building the unwind table. Within each function, DWARF maintains a
\emph{canonical frame address} (CFA), which is invariant in the function body,
and is traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which has been validated with the DWARF
unwind table validator tool~\cite{Bastian}. When starting to run on a new
fiber, we insert DWARF bytecode to use the follow the |parent_fiber| pointer
and dereference the |saved_sp| to get the CFA. During callback, we save the
current system stack pointer in the DWARF block in Figure~\ref{fig:fiber} to
identify the CFA in the C stack. DWARF unwinding for external calls is
implemented similarly by saving the current OCaml stack pointer. With these
changes, we get the same backtrace for the |meander.exe| program from
\S\ref{sec:unwind}, modulo runtime system functions due to effect handlers.
